Type Visualization for R

### 2020-06-21 paper/main.tex
paper: More edits

\section{Data and Collection}

@@

Execution traces were recorded by running a dynamic analysis
over the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of function type signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, the function call {\tt !(FALSE)}
would be recorded as the type signature {\tt logical $\to$ logical},
meaning that the function {\tt !} took a logical value
and returned a logical value.\footnote{
  In R, a logical value is a boolean.
}
The type tags are assigned by the dynamic analysis
and refine the actual type tags in R.
For instance, the analysis
assigns the scalar tag {\tt double} to a singleton vector of double values
because R does not have native support for scalars.
The data set is a table of reduced execution traces.
Because \typevis was designed for the analysis of types,
not all information in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

@@

\begin{enumerate}
\item {\bf Find Function.} Quickly search for a particular function and identify basic information. This includes how often the function is called and which package contains it.
\item {\bf Determine Types.} Given a specific function, determine the argument types, return types, and identify the frequency of specific type signatures. A user should be able to readily infer, for example, if a function is polymorphic or if it is a predicate.
\item {\bf Compare Signatures.} Compare the occurrence of type signatures within a specific function, and across many different functions. Users must be able to isolate individual signatures and see them in the context of others.
\end{enumerate}

@@

\typevis is an interactive web-based visualization
that permits real-time exploration,
even with large data sets.
The default data set contains over a million rows
based on the analysis of almost five hundred
different R packages.
Figure~\ref{fig:typevis} displays
the heart of \typevis
in its initial configuration.
Most aspects of the system can be tweaked
by a user if desired.

@@

The right-hand panel of figure~\ref{fig:typevis}
contains three elements: a search bar,
a package treemap,
and a function treemap.
All three are linked.
Selecting a package
in the top treemap will immediately narrow the
bottom view to only functions available in that package.
Additionally, entering a function name in the search
will select it in the function treemap.

These different components support different modes of search.
If a user has a specific function in mind,
the search bar supports autocomplete
that can quickly select a function among thousands.
However, if one wants to browse
all available functions and packages,
the treemap view is more applicable.
Nodes in the bottom treemap convey the log-scaled
frequency of function calls,
and correspondingly the top treemap
is scaled based on invocations
of functions defined in that package.
Log-scaling and pagination keep
elements of the treemap legible,
even if absolute size differences are significant.

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the data set was more deeply nested,
\typevis can be configured to
handle that.

If configured, \typevis allows for multiple function selection.
Both the {\tt length.unit} and {\tt valid.unit} functions
from the {\tt grid} package are selected in figure~\ref{fig:typevis}.
These functions are rendered simultaneously in the type flow
panel, allowing one to make comparisons between them.

@@

Prominently featured as the center visualization,
a parallel sets diagram encodes type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara et al.~\cite{kosara:2006},
are ideally suited for type signatures.
Each data point is can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n)$
corresponding to the function name,
argument types,
and return type.
Every component is categorical,
and we are interested in visualizing frequencies of these points.
This is the exact use-case for parallel sets.

Flows begin at nodes labeling a function,
curve across each argument type in order,
and terminate at the return type.
Widths are proportional to how many times a function
is called with that signature during analysis.
Flow segments are filled with a hue-varying color scale,
based on the type where the segment ends.
Beneath each node label is an approximate count.
If the node represents a function, then the
number is how many times the function was invoked.
If the node represents a type, then the
number is how often a value of that type
at that argument position was recorded during analysis.

For example, in figure~\ref{fig:typevis},
the {\tt length.unit} function is called with
two different type signatures at run-time.
Most frequently, the function is used with the
signature {\tt double $\to$ integer},
but a substantial fraction of the time it is
used as {\tt double[] $\to$ integer}.
Here, {\tt double[]} stands for an array of
doubles with arbitrary dimension.

Interactivity significantly augments the
usability of the parallel sets visualization.
Hovering over a flow will highlight the
entire path of the flow
and will fade other functions
into the background.
When displaying many flows at the same time,
highlighting becomes especially critical
for user comprehension.
Additional quantitative information
about the flow is also supplied
while hovering.
Clicking a flow will focus on that function
only, filtering away all others.

Flow layout is a critical consideration
for a sufficiently legible visualization.
Figure~\ref{fig:decross} exemplifies what
happens if flow layout is not adequately handled.
The top layout was generated without any
flow decrossing algorithm,
while the bottom layout does
apply decrossing.
Even with highlighting,
the utility of the visualization is vastly
reduced without some decrossing effort.
To maintain real-time performance,
we use two different decrossing methods.
First, the system attempts to construct
a layout with the globally optimal
minimum amount of flow crossing by
solving a mixed-integer linear program.
Unfortunately, for even moderately
sized graphs this can take too long to compute.
Therefore, after about $1$ second
the optimal algorithm is terminated
if it has not finished, and a faster
algorithm that only locally minimizes crossing
between layers computes the layout.

@@

Since the amount of data that a user may query is large,
the application consists of both a frontend and a backend.

On the frontend, state management, interactivity,
and DOM manipulation is all done within the Vue framework.
This is in contrast with many web-based visualizations
that use D3 for DOM manipulation.
We chose this setup because of the many different
individual components whose state must be kept
synchronized.
This includes not only the visualizations, but also
user interface elements like the autocomplete search bar.
In our experience, Vue's model of reactivity is
more successful at maintaining consistency across the
entire system than using D3 and vanilla JavaScript.
While \typevis does not use D3 for DOM manipulation,
it still does use D3 for other visualization-related
computations.

Our backend is a Node.js application that serves
JSON in a standard REST-style architecture.
Data is stored in a SQLite database that has been
indexed to achieve lookups as fast as possible.
To avoid excessive memory consumption and speed up computations,
clients query data from the server on an as-needed basis.
Thus, a stable network connection is required for
real-time interaction.

@@

\typevis suffers from a number of shortcomings
and does not fully address the entire range of possible tasks
a language designer may need to perform.

In particular, R supports a wide range of data types
including types that contain dimensionality information.
For example, an R array may have type {\tt integer[3]}
indicating that it is an array of length $3$.
This will get compressed down into a single array
type labeled {\tt integer[]}.
Reducing the number of types can make the visualization
more understandable---at the expensive of losing some
precision in type information.
Ideally a user would be able to interactively tune
the level of type granularity based on their interest
in a specific property.
Such a feature would be especially helpful for R
since values can come attached with relevant additional
data such as the dynamic dispatch method
or user-specified metadata.

Filtering is an central mechanism in the system,
but it also comes at a cost.
If one has a specific function or several functions
in mind, then \typevis provides a useful
local view of that type information.
However, it does not support any kind of global
view of type information across a large amount of
functions. When the number of flows or nodes
becomes too great, the visualization will start
paginating, making comparisons over large amounts
of data impossible.
Aggregation and summarization would be key to
making a global view of the data feasible---that
remains future work.

@@

We have presented \typevis, an interactive visualization
of run-time type information for R.
It supports real-time exploration of a data set of
execution traces from a large corpus of R packages.
To narrow the data under consideration,
\typevis provides facilities for filtering
and searching.
Type signatures are represented as flows traversing across
each argument type and return type in sequence.
Flow widths are proportional to how often
that type signature was seen during the dynamic analysis.
Interactions provide additional details and
assist users in comprehension and navigation.

Our tool is aimed at elucidating how R programmers
use the language in practice.
Specifically, the data and visualization will be used to
inform the design of a gradual type system for R.
However, \typevis is not tied to the R ecosystem
and is suitable to use for any dynamically typed language
with a similar data set.
Additionally, the visualization could yield insights
useful for other purposes than gradual typing.

We hope that, in the future, programming language designers
will collect and use empirical data about how their languages
are used in practice.
Such data can be used to make informed decisions about
language design.
Visualization will be unavoidable in these scenarios
since semantically-meaning information about programs
is rich in structure.
Generic, canned visualizations will not be sufficient.
\typevis is just the first step in this direction.

\acknowledgments{
  The authors wish to thank the students of CS 7250 for useful design feedback,
  and contributors to all the open source projects that \typevis depends on.
  This work was supported in part by a Northeastern Graduate Fellowship.
}

>>>>

\section{Data and Collection} \label{sec:data}

@@

Execution traces were recorded by running a dynamic analysis
over the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of function type signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, the function call {\tt !(FALSE)}
would be recorded as the type signature {\tt logical $\to$ logical},
meaning that the function {\tt !} took a logical value (i.e. a boolean)
and returned a logical value.
The type tags are assigned by the dynamic analysis
and refine the actual type tags in R.
For instance, the analysis
assigns the scalar tag {\tt double} to a singleton vector of double values
because R does not have native support for scalars.
The data set is a table of reduced execution traces.
Because \typevis was designed for the analysis of types,
not all information in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

@@

\begin{enumerate}
\item {\bf Find Function.} Quickly search for a particular function and identify basic information. This includes how often the function is called and which package contains it.
\item {\bf Determine Types.} Given a specific function,
  determine its argument and return types, and identify the frequency of specific type signatures.
  A user should be able to readily infer, for example,
  if a function is polymorphic (i.e. accepts arguments of multiple different types) or if it is a predicate (i.e. returns a logical value).
\item {\bf Compare Signatures.} Compare the occurrence of type signatures within a specific function, and across many different functions. A user must be able to isolate individual signatures and exmaine them in the context of others.
\end{enumerate}

@@

\typevis is an interactive web-based visualization
that permits real-time exploration,
even with large data sets.
The default data set,
described in \S\ref{sec:data},
contains over a million rows.
Figure~\ref{fig:typevis} shows
the heart of \typevis
in its initial configuration.
Most aspects of the system can be tweaked
by the user if desired.

@@

The right-hand panel of figure~\ref{fig:typevis}
contains three elements: a search bar,
a package treemap,
and a function treemap.
All three are linked.
Selecting a package
in the top treemap will cause the
bottom view to show only functions available in that package.
Entering a function name in the search bar
will select it in the function treemap.

These different components support different modes of search.
If a user has a specific function in mind,
the search bar supports autocomplete
that can select one function among thousands.
However, if one wants to browse
all available functions and packages,
the treemap view is more applicable.
Area in the bottom treemap conveys the log-scaled
frequency of function calls,
and correspondingly the top treemap
is scaled based on invocations
of functions defined in that package.
Log-scaling and pagination keep
elements of the treemap legible,
even when absolute size differences are significant.

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
The packages in our data are not deeply structured,
so the treemaps in figure~\ref{fig:typevis} are only one level deep.
However, if the data set recorded deeply nested packages,
\typevis could be modified to handle that.

If configured, \typevis allows for multiple function selection.
In this example, the {\tt length.unit} and {\tt valid.unit} functions
from the {\tt grid} package are selected.
These functions are rendered simultaneously in the type flow
panel, allowing the user to make comparisons between them.

@@

Prominently featured in the center of figure~\ref{fig:typevis}
is a parallel sets diagram encoding the type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara et al.~\cite{kosara:2006},
are ideally suited for type signatures.
Each data point can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n)$
corresponding to the function name,
argument types,
and return type.
Every component of that tuple is categorical,
and we are interested in visualizing the frequencies of these points.
This is the exact use-case for parallel sets.

Flows begin at a node labeling a function,
curve across each argument type in order,
and terminate at the return type.
Widths are proportional to how many times a function
is called with that signature during analysis.
Flow segments are filled with a hue-varying color scale,
determined by the type where the segment ends.
Beneath each node label is an approximate count of either:
how many times the function was invoked if the node represents a function,
or how often a value of that type was recorded during analysis if the node represents a tuple.

In figure~\ref{fig:typevis},
the {\tt length.unit} function is called with
two different type signatures at run-time.
Most frequently, the function is used with the
signature {\tt double $\to$ integer},
but a substantial fraction of the time it is
used as {\tt double[] $\to$ integer}.
Here, {\tt double[]} stands for an array of
doubles with arbitrary dimension.

Interactivity significantly enhances the
usability of the parallel sets diagram.
Hovering over a flow will highlight the
entire path of the flow
and will fade out all other functions.
When displaying many flows at the same time,
highlighting becomes critical
for user comprehension.
Additional quantitative information
about the flow is supplied on-demand while hovering.
Clicking a flow will focus on that function
only, filtering away all others.

Flow layout is a necessary consideration
for a legible visualization~\cite{sugiyama:1981}.
Figure~\ref{fig:decross} exemplifies what
happens if flow layout is not adequately computed.
The top layout was generated without any
flow decrossing algorithm,
while the bottom layout applies optimal decrossing.
Even with highlighting,
the visualization is almost impossible to understand
due to excessive flow overlap.
To maintain real-time performance,
we use two different decrossing methods.
First, the system attempts to construct
a layout with the globally optimal solution,
minimizing the amount of flow crossing by
solving a mixed-integer linear program.
Unfortunately, for even moderately
sized graphs this can take too long to compute.
Therefore, after about $1$ second,
the optimal algorithm is terminated
if it has not finished, and a faster
algorithm that only minimizes local crossing between layers
computes the layout.
While not ideal, it is much better than no decrossing
at all.

@@

Since the amount of data a user may query is large,
the application consists of both a frontend and a backend.
This avoids excessive memory consumption and speeds up computations,
as the client queries data from the server on an need-to-know basis.

On the frontend, state management, interactivity,
and DOM manipulation is all done with the Vue framework.
This differs from many web-based visualizations
that use D3 for DOM manipulation.
We chose this setup because \typevis retains many distinct
components whose state must be kept synchronized.
This includes not only the visualizations, but also
user interface elements like the search bar.
In our experience, Vue's model of reactivity is
more successful at maintaining consistency across the
entire application than D3 and vanilla JavaScript.
While \typevis does not use D3 for DOM manipulation,
it still does use it for other visualization-related
tasks. Decrossing is computed using web workers
to allow early termination if the thread takes too long.

Our backend is a Node.js application that serves
JSON with a standard REST-style architecture.
Data is stored in a SQLite database that has been
indexed to achieve lookups as fast as possible.
A stable network connection is required for
real-time interaction.

@@

\typevis suffers from some shortcomings
and does not fully address the entire range of tasks
a language designer may need to perform.

In particular, R supports a range of data types
including ones containing dimensionality information.
For example, an R array may have type {\tt integer[3]}
indicating that it is an array of length $3$.
This will get compressed down into a single array of
type {\tt integer[]}.
Reducing the number of types can make the visualization
more understandable, at the expensive of losing
precision in type information.
Ideally a user would be able to interactively tune
the level of granularity based on their interest.
Such a feature would be especially helpful for R
since values can come attached with lots of relevant
data such as the dynamic dispatch method
or user-specified metadata.

Filtering is a central mechanism of our system,
but it also comes at a cost.
If one has a specific function or several functions
in mind, then \typevis provides a useful
local view of that information.
It does not support, however, any kind of global
view of type information across a large number of
functions. When the amount of flows or nodes
becomes too great, the visualization will start
paginating, making comparisons across all flows impossible.
Aggregation and summarization would be key to
making a global view of the data feasible, but that remains future work.

@@

We have presented \typevis, an interactive visualization
of run-time type information
that supports real-time analysis and exploration.
The tool was developed for a massive data set of
execution traces from a corpus of popular R packages.

\typevis provides facilities for filtering, searching,
understanding, and comparing type information.
Signatures are represented as flows traversing
each argument type and return type in sequence.
Interactions provide additional details and
assist users with comprehension and navigation.

Our system is aimed at elucidating how R programmers
use the language in the field.
Specifically, the data and visualization will be used to
inform the design of a future gradual type system for R.
However, \typevis is not tied to the R ecosystem
and is suitable to use for any dynamically-typed language.
The visualization could yield insights
useful for purposes other than gradual typing,
and beyond what we have considered.

We hope that, in the future, programming language designers
will collect and use empirical data about their languages.
Such data could be used to make informed decisions about design,
where the cost of making a mistake is high.
Visualization will be unavoidable in these scenarios
since semantically-meaning information about programs
is rich in structure.
Generic, canned visualizations will not be sufficient.
\typevis is a first step in this direction.

\acknowledgments{
  The authors wish to thank
  Alexi Turcotte,
  Cody Dunne,
  Sara Di Bartolomeo,
  Younes El Idrissi Yazami,
  Peta Maj,
  the students of CS 7250,
  and contributors to all the open source projects that \typevis uses.
  This work was supported in part by a Northeastern Graduate Fellowship.
}
